Dependently Typed Languages in Statix

### 2022-10-14 paper-evcs/oasics-v2021-sample-article.tex
Minor changes

\begin{abstract}
When working with the Spoofax workbench, the Statix meta-language can be used for the specification of Static Semantics. Statix aims to cover a broad range of languages and type-systems. However, no attempts have been made to express dependently typed languages in Statix. Type-checking dependent languages is challenging since there is a need to evaluate terms during type-checking. This paper presents how to make a dependent language by implementing the calculus of constructions in Statix. 
\end{abstract}

@@

Spoofax is a textual language workbench: \label{key}a collection of tools that enable the development of textual languages\cite{spoofax}. When working with the Spoofax workbench, the Statix meta-language can be used for the specification of Static Semantics. To provide these advantages to as many language developers as possible, Statix aims to cover a broad range of languages and type-systems. However, no attempts have been made to express dependently typed languages in Statix.

@@

The implementation of the language is available on github. The \verb|final-simple| branch contains the implementation that is relevant for this paper. \\
\url{https://github.com/JonathanBrouwer/master-thesis/}

@@

The base language that was implemented is the Calculus of Constructions \cite{Coquand_Huet_1988}, the language at the top of the lambda cube \cite{lambda_cube}. One extra feature was added that is not present in the Calculus of Constructions, let bindings. Let bindings could be desugared by substituting, but this may grow the program size exponentially, so having them in the language is useful. The abstract syntax of the language is available in figure \ref{fig:syntax}.

@@

Both of these need to be stored in the scope graph. The scope graph only has a single type of edge, called \verb|P| (parent) edges. It also only has a single relation, called \verb|name|. This name stores a \verb|NameEntry|, which can be either a \verb|NType|, which stores the type of a name, or a \verb|NSubst|, which stores a substitution corresponding to a name.

\begin{lstlisting}
signature sorts	NameEntry
constructors    NType : Expr -> NameEntry
                NSubst : scope * Expr -> NameEntry
relations       name : ID -> NameEntry
name-resolution	labels P
\end{lstlisting}

These are all the definitions we will need to type-check programs. Next, we will introduce some Statix relations that can be used to interact with these scope graphs:

\begin{lstlisting}
sPutType  : scope * ID * Expr -> scope
sPutSubst : scope * ID * (scope * Expr) -> scope
sGetName  : scope * ID -> NameEntry
sGetNames : scope * ID -> list((path * (ID * NameEntry)))
sEmpty    : -> scope
\end{lstlisting}

The \verb|sPutType| and \verb|sPutSubsts| relations generate a new scope given a parent scope and a type or a substitution respectively. To query the scope graph, use \verb|sGetName| or \verb|sGetNames|, which will return a \verb|NameEntry| or a list of \verb|NameEntries| respectively that the query found. Finally, \verb|sEmpty| returns a fresh empty scope.

\subsection{Beta Reductions}

@@

We need to define multiple relations that will be used later for type-checking. First, the primary relation is \verb|betaReduceHead|, that takes a scoped expression and a stack of applications, and returns a head-normal expression. The scope acts as the environment from \cite{krivine} paper, using \verb|NSubst| to store substitutions. All rules for \verb|betaReduceHead| are given in figure \ref{fig:beta-head-reduce-rules}. We use the syntax $s1 \vdash e1 \bhr s2 \vdash e2$ to express \verb|betaReduceHead((s1, e1)) == (s2, e2)|. Figure \ref{fig:beta-head-reduce-rules} contains the rules necessary for beta head reduction of the language. One relation that is used for this is the \verb|rebuild| relation, which takes a scoped expression and a list of arguments and converts it to an expression by adding \verb|FnDestruct|s.

@@

		\inferrule{ 
		}{s \vdash Let(n, v, b), [] \bhr sPutSubst(s, n, (s, v)) \vdash b }

>>>>

\begin{abstract}
When working with the Spoofax workbench, the Statix meta-language can be used for the specification of static semantics. Statix aims to cover a broad range of languages and type-systems. However, no attempts have been made to express dependently typed languages in Statix. Type-checking dependent languages is challenging since there is a need to evaluate terms during type-checking. This paper presents how to make a dependent language by implementing the calculus of constructions in Statix. 
\end{abstract}

@@

Spoofax is a textual language workbench: \label{key}a collection of tools that enable the development of textual languages\cite{spoofax}. When working with the Spoofax workbench, the Statix meta-language can be used for the specification of static semantics. To provide these advantages to as many language developers as possible, Statix aims to cover a broad range of languages and type-systems. However, no attempts have been made to express dependently typed languages in Statix.

@@

The implementation of the language is available on GitHub. The \verb|final-simple| branch contains the implementation that is relevant for this paper. \\
\url{https://github.com/JonathanBrouwer/master-thesis/}

@@

The base language that has been implemented is the Calculus of Constructions \cite{Coquand_Huet_1988}, the language at the top of the lambda cube \cite{lambda_cube}. One extra feature was added that is not present in the Calculus of Constructions, let bindings. Let bindings could be desugared by substituting, but this may grow the program size exponentially, so having them in the language is useful. The abstract syntax of the language is available in figure \ref{fig:syntax}.

@@

Both of these need to be stored in the scope graph. The scope graph only has a single type of edge, called \verb|P| (parent) edges. It also only has a single relation, called \verb|name|. This name stores a \verb|NameEntry|, which can be either a \verb|NType|, which stores the type of a name, or a \verb|NSubst|, which stores a name that has been substituted with a value.

Next, we will introduce some Statix relations that can be used to interact with these scope graphs:

\begin{lstlisting}
sPutType  : scope * ID * Expr -> scope
sPutSubst : scope * ID * (scope * Expr) -> scope
sGetName  : scope * ID -> NameEntry
sGetNames : scope * ID -> list((path * (ID * NameEntry)))
sEmpty    : -> scope
\end{lstlisting}
The \verb|sPutType| and \verb|sPutSubsts| relations generate a new scope given a parent scope and a type or a substitution respectively. To query the scope graph, use \verb|sGetName| or \verb|sGetNames|, which will return a \verb|NameEntry| or a list of \verb|NameEntries| respectively that the query found. Finally, \verb|sEmpty| returns a fresh empty scope.

\newpage
\subsection{Beta Reductions}

@@

We need to define multiple relations that will be used later for type-checking. First, the primary relation is \verb|betaReduceHead|, that takes a scoped expression and a stack of applications, and returns a head-normal expression. The scope acts as the environment from \cite{krivine} paper, using \verb|NSubst| to store substitutions. All rules for \verb|betaReduceHead| are given in figure \ref{fig:beta-head-reduce-rules}. We use the syntax $s1 \vdash e1, t \bhr s2 \vdash e2$ to express \verb|betaReduceHead((s1, e1)) == (s2, e2)|. Figure \ref{fig:beta-head-reduce-rules} contains the rules necessary for beta head reduction of the language. One relation that is used for this is the \verb|rebuild| relation, which takes a scoped expression and a list of arguments and converts it to an expression by adding \verb|FnDestruct|s.

@@

		\inferrule{ 
			sPutSubst(s, n, (s, v)) \vdash b, t \bhr s' \vdash e'
		}{s \vdash Let(n, v, b), t \bhr s' \vdash e' }
